---
title: "Bray-Curtis sample-to-sample and Shannon diversity, country and T1D status"
output: html_notebook
---

load packages

```{r}
library(ggplot2)
library(data.table)
library(stringr)
library(dplyr)
library(tidyr)
library(rprojroot)
library(vegan)
library(ggpubr)
library(rstatix)
library(cowplot)
library(emmeans)
```

set paths and filenames

```{r}
### files
long_table=sprintf("%s/data/mp142_TGVG1.1_MPA4_combined_abundance_table_longform1.tsv",
                   find_rstudio_root_file())
metadata_table=sprintf("%s/data/some_teddy_MP142_metadata2.all_samples1.delivery.csv",
                       find_rstudio_root_file())

T1D_groups_file=sprintf("%s/data/MP142_CASE_CNTRL_T1D_LIST1.csv", 
                        find_rstudio_root_file())

```

load long table and metadata, merge

```{r}
long_dt <- fread(sprintf("%s", long_table), sep = "\t", header = T) %>%
  select(sampleID, rel_abundance, lineage) %>%
  mutate(sampleID = as.character(sampleID),
         kingdom = case_when(grepl("k__Bac", lineage) ~ "Bacteria", 
                             grepl("k__Vir", lineage) ~ "Virus",
                             grepl("k__Ar", lineage) ~ "Archea",
                             grepl("k__Euk", lineage) ~ "Eukaryota",
                             TRUE ~ "other"))

groups_dt <- fread(T1D_groups_file, sep = ",", header = T) %>%
  select(c(mask_id, case_ind))

meta_dt <- fread(sprintf("%s", metadata_table), sep = ",", header = T) %>%
  select(-V1) %>%
  mutate(sample = as.character(sample))

merge_full_dt <- merge(long_dt, meta_dt, by.x = "sampleID", by.y ="sample")

## filtering down to only subject from TEDDY T1D groups

merge_dt <- merge(merge_full_dt, groups_dt, by = "mask_id")

rm(merge_full_dt)
```

wide format, bray-curtis, VIROME

```{r}

mask_id_list <- as.list(
  merge_dt %>% distinct(mask_id)
)

big_bray_long <- data.table(sample_info1=character(), 
                            sample_info2=character(), 
                            value=numeric())

## big for loop to do each subject
for (subjectq in mask_id_list[1]$mask_id) {
  temp_long_dt <- merge_dt %>%
    filter(mask_id == subjectq)
  
  temp_long_dt$sample_info <- str_c(temp_long_dt$sampleID, "@", 
                                    temp_long_dt$mask_id, "@", 
                                    temp_long_dt$age_days)
  
  temp_long_dt <- temp_long_dt %>%
    select(c(sample_info, rel_abundance, lineage, kingdom)) %>%
    distinct()
  if(nrow(temp_long_dt) > 1){
    wide_virome_temp_dt <- temp_long_dt %>%
      filter(kingdom == "Virus") %>%
      select(-kingdom) %>%
        pivot_wider(names_from = lineage, 
                  values_from = rel_abundance, 
                  values_fill = 0)
    sample_info_l <- wide_virome_temp_dt$sample_info
  
    wide_virome_temp_dt <- wide_virome_temp_dt %>% select(-sample_info)
    
    bray_curtis_dist <- vegdist(wide_virome_temp_dt, method="bray")
    
    bray_curtis_mat <- as.matrix(bray_curtis_dist)
    
    bray_curtis_df <- as.data.frame(bray_curtis_mat)
    
    colnames(bray_curtis_df) <- sample_info_l
    rownames(bray_curtis_df) <- sample_info_l
    bray_curtis_df$sample_info1 <- rownames(bray_curtis_df)
  
    bray_long <- melt(setDT(bray_curtis_df), 
                      id.vars = c("sample_info1"), 
                      variable.name = "sample_info2") %>%
      filter(sample_info1 != sample_info2)
    big_bray_long <- rbind(big_bray_long, bray_long)
  }
}

big_bray_long <- big_bray_long[, 
                               c("sampleID1", "mask_id1", "day_of_life1") 
                               := tstrsplit(sample_info1, "@", fixed=TRUE)]

big_bray_long <- big_bray_long[, 
                               c("sampleID2", "mask_id2", "day_of_life2") 
                               := tstrsplit(sample_info2, "@", fixed=TRUE)] %>%
  select(c(sampleID1, sampleID2, mask_id1, day_of_life1, day_of_life2, value)) %>%
  mutate(days_apart = as.numeric(day_of_life2) - as.numeric(day_of_life1)) %>%
  filter(days_apart >= 1) 
  
big_bray_long <- big_bray_long %>%
  rename(bray_curtis = value)
```

quick plot of days apart vs bray-curtis distance, virome

```{r}
big_bray_long %>%
  ggplot(aes(x= days_apart, y = bray_curtis)) +
  geom_point(alpha = 0.005, color = "cadetblue") +
  geom_smooth() +
  theme_bw()
```


```{r}
merge(big_bray_long, meta_dt, 
      by.x = "sampleID1", by.y = "sample") %>%
  ggplot(aes(x= days_apart, y = bray_curtis, color = T1D)) +
  geom_point(alpha = 0.005) +
  geom_smooth() +
  theme_bw()
  
  
```


get bray-curtis distance to next sample VIROME

```{r}
vir_next_bray_dt <- big_bray_long %>%
  group_by(mask_id1) %>%
  filter(n_distinct(sampleID1) >= 2) %>%
  ungroup() %>%
  group_by(sampleID1) %>%
  filter(days_apart == min(days_apart))

vir_next_bray_dt <- merge(vir_next_bray_dt, meta_dt, 
                          by.x = "sampleID1", by.y = "sample")

## add groups
vir_next_bray_dt <- merge(vir_next_bray_dt, groups_dt, by = "mask_id")

```

```{r}

t1d_lm <- lm(days_apart ~ bray_curtis*T1D, 
             data= vir_next_bray_dt %>%
               filter(days_apart <= 100,
                      days_apart >= 10))

( fitted.emt <- emtrends(t1d_lm, "T1D", var = "bray_curtis") )

pw_t1d <- pairs(fitted.emt)

pw_t1d
```


```{r}
vir_days_p <- vir_next_bray_dt %>%
  filter(days_apart <= 100,
         days_apart >=10) %>%
  ggplot(aes(x= days_apart, y = bray_curtis, color = T1D)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm"
    ) +
  geom_text(y = 0.9,
            x = 50,
            label = "p-value = 0.0234",
            color = "black") +
  scale_color_manual(values = c("#8CBEB1", "orangered")) +
  theme_bw()  +
  labs(x = "Days to Next Sample",
       y = "Bray-Curtis Dissimilarity")

vir_days_p

ggsave(
  vir_days_p,
  file = sprintf(
    "%s/charts/bray_curtis_next_vir_days1.pdf",
    find_rstudio_root_file()
  ),
  height = 5, width = 3.5
)
```



average bray distance by subject/pair VIROME
```{r}
vir_bray_sum_subj <- vir_next_bray_dt %>%
  group_by(mask_id1) %>%
  summarize(average_bray = mean(bray_curtis),
            T1D = T1D,
            pair_num = case_ind) %>%
  ungroup() %>%
  distinct() %>%
  ## I think there is a typo in group 114
  filter(!pair_num == 114)
```

```{r}


vir_bray_subj_wide <- vir_bray_sum_subj %>%
  group_by(pair_num) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  pivot_wider(id_cols = pair_num, names_from = T1D, values_from = average_bray)
```

```{r}
vir_bray_subj_wide %>%
  ggpaired(cond1 = "Yes", cond2 = "No",
    fill = "condition", palette = c("#8CBEB1", "orangered"),
    line.color = "lightgrey", ggtheme = theme_bw(), line.size = 0.1) +
  stat_compare_means(paired = TRUE) 
  
```

paired wilcox test
```{r}
wilcox.test(vir_bray_subj_wide$Yes, vir_bray_subj_wide$No, paired = TRUE)

```

```{r}
vir_bray_sum_subj %>%
  group_by(pair_num) %>%
  filter(n() != 2)
```

new plot average bray distance by T1D and country status VIROME
```{r}

vir_bray_sum_subj  %>%
  group_by(pair_num) %>%
  filter(n() == 2) %>%
  ggplot(aes(T1D, average_bray)) +
  geom_boxplot(aes(fill = T1D), outlier.shape = NA, alpha = 0.6) +
  geom_point(color="grey20", alpha = 0.5, stroke = 0) +
  geom_line(aes(group = pair_num), color = "lightgrey", alpha = 0.5) +
  geom_text(label = "Wilcoxon,\np-value = 0.015", x = "No", 
            y = 1, size = 8/.pt) +
  scale_fill_manual(values=c("#8CBEB1", "orangered")) +
  ylim(c(0.2, 1.05)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_blank()) +
  labs(y = "Bray-Curtis Distance\nto Follow-up Sample",
       title="virus SGBs")

ggsave(file = sprintf("%s/charts/virome_bray_curtis_T1D_NCC_paired.pdf",
                      find_rstudio_root_file()), 
                 width = 2.5, height = 3.5)  
```

```{r}
vir_bray_sum_subj  %>%
  group_by(pair_num) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  group_by(T1D) %>%
  summarize(avg_bray_t1d = mean(average_bray),
            n = n_distinct(mask_id1))
```





```{r}
## this functions rounds a number up to the nearest 100
round_any = function(x, accuracy, f=ceiling){f(x/ accuracy) * accuracy}

DOL_vir_pair_dt <- vir_next_bray_dt %>%
  mutate(rounded_DOL =  round_any(age_days, 100)) %>%
  filter(rounded_DOL <= 1400) %>%
  group_by(mask_id1, rounded_DOL) %>%
  summarize(average_bray = mean(bray_curtis),
            T1D = T1D,
            pair_num = case_ind) %>%
  ungroup() %>%
  distinct() %>%
  filter(!pair_num == 114)
```

```{r}
DOL_vir_pair_subj_wide <- DOL_vir_pair_dt %>%
  group_by(pair_num, rounded_DOL) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(pair_num, rounded_DOL), names_from = T1D, 
              values_from = average_bray)
```

```{r}
DOL_vir_pair_subj_wide %>%
  ggpaired(cond1 = "Yes", cond2 = "No",
    fill = "condition", palette = c("#8CBEB1", "orangered"),
    line.color = "lightgrey", ggtheme = theme_bw(), line.size = 0.1,
    facet.by = "rounded_DOL") +
  stat_compare_means(paired = TRUE) 
```

```{r}
DOL_vir_pair_wilcox_dt <- DOL_vir_pair_subj_wide %>% 
  group_by(rounded_DOL) %>%
  do(w = wilcox.test(.$Yes, .$No, data=., paired=TRUE)) %>% 
       summarise(rounded_DOL, 
                 Wilcox = w$p.value,
                 aster = case_when(Wilcox < 0.0001 ~ "****",
                                   Wilcox < 0.001 ~ "***", 
                                   Wilcox < 0.01 ~ "**",
                                   Wilcox < 0.05 ~ "*",
                                   TRUE ~ "ns"))

DOL_vir_pair_wil_dt <- merge(DOL_vir_pair_dt, DOL_vir_pair_wilcox_dt, by = "rounded_DOL")
```

```{r}
DOL_vir_pair_wil_dt %>%
  group_by(rounded_DOL,pair_num) %>%
  filter(n() == 2) %>%
  ggplot(aes(T1D, average_bray)) +
  geom_boxplot(aes(fill = T1D), outlier.shape = NA, alpha = 0.6) +
  geom_point(color="grey20", alpha = 0.5, stroke = 0, size = 0.8) +
  geom_line(aes(group = pair_num), color = "lightgrey", alpha = 0.3) +
  geom_text(data = DOL_vir_pair_wilcox_dt, aes(label = aster), x = "No", y = 1.05) +
  scale_fill_manual(values=c("#8CBEB1", "orangered")) +
  ylim(c(NA, 1.1)) +
  facet_wrap(vars(rounded_DOL), nrow = 1) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_blank()) +
  labs(x = "day of life (rounded)", 
       y = "Bray-Curtis Distance\nto Follow-up Sample",
       title="virus SGBs")

ggsave(file = sprintf("%s/charts/virome_T1D_and_dayoflife_bray_curtis_NCC_paired.pdf",
                      find_rstudio_root_file()), 
                 width = 6, height = 3.5)  
```




intrasubject bray-curtis, BACTERIOME

```{r}

mask_id_list <- as.list(
  merge_dt %>% distinct(mask_id)
)

bac_bray_long <- data.table(sample_info1=character(), 
                            sample_info2=character(), 
                            value=numeric())

for (subjectq in mask_id_list[1]$mask_id) {
  temp_long_dt <- merge_dt %>%
    filter(mask_id == subjectq)
  
  temp_long_dt$sample_info <- str_c(temp_long_dt$sampleID, "@", 
                                    temp_long_dt$mask_id, "@", 
                                    temp_long_dt$age_days)
  
  temp_long_dt <- temp_long_dt %>%
    select(c(sample_info, rel_abundance, lineage, kingdom)) %>%
    distinct()
  if(nrow(temp_long_dt) > 1){
    wide_virome_temp_dt <- temp_long_dt %>%
      filter(kingdom == "Bacteria") %>%
      select(-kingdom) %>%
        pivot_wider(names_from = lineage, 
                  values_from = rel_abundance, 
                  values_fill = 0)
    sample_info_l <- wide_virome_temp_dt$sample_info
  
    wide_virome_temp_dt <- wide_virome_temp_dt %>% select(-sample_info)
    
    bray_curtis_dist <- vegdist(wide_virome_temp_dt, method="bray")
    
    bray_curtis_mat <- as.matrix(bray_curtis_dist)
    
    bray_curtis_df <- as.data.frame(bray_curtis_mat)
    
    colnames(bray_curtis_df) <- sample_info_l
    rownames(bray_curtis_df) <- sample_info_l
    bray_curtis_df$sample_info1 <- rownames(bray_curtis_df)
  
    bray_long <- melt(setDT(bray_curtis_df), 
                      id.vars = c("sample_info1"), 
                      variable.name = "sample_info2") %>%
      filter(sample_info1 != sample_info2)
    bac_bray_long <- rbind(bac_bray_long, bray_long)
  }
}

bac_bray_long <- bac_bray_long[, 
                               c("sampleID1", "mask_id1", "day_of_life1") 
                               := tstrsplit(sample_info1, "@", fixed=TRUE)]

bac_bray_long <- bac_bray_long[, 
                               c("sampleID2", "mask_id2", "day_of_life2") 
                               := tstrsplit(sample_info2, "@", fixed=TRUE)] %>%
  select(c(sampleID1, sampleID2, mask_id1, day_of_life1, day_of_life2, value)) %>%
  mutate(days_apart = as.numeric(day_of_life2) - as.numeric(day_of_life1)) %>%
  filter(days_apart >= 1) 
  
bac_bray_long <- bac_bray_long %>%
  rename(bray_curtis = value)
```

quick plot of days apart vs bray-curtis distance, bacteriome

```{r}
bac_bray_long %>%
  ggplot(aes(x= days_apart, y = bray_curtis)) +
  geom_point(alpha = 0.005, color = "cadetblue") +
  geom_smooth() +
  theme_bw()
```

get bray-curtis distance to next sample BACTERIOME

```{r}
bac_next_bray_dt <- bac_bray_long %>%
  group_by(mask_id1) %>%
  filter(n_distinct(sampleID1) >= 2) %>%
  ungroup() %>%
  group_by(sampleID1) %>%
  filter(days_apart == min(days_apart))

bac_next_bray_dt <- merge(bac_next_bray_dt, meta_dt, by.x = "sampleID1", by.y = "sample")

## add groups
bac_next_bray_dt <- merge(bac_next_bray_dt, groups_dt, by = "mask_id")

```


```{r}

bac_t1d_lm <- lm(days_apart ~ bray_curtis*T1D, 
             data= bac_next_bray_dt %>%
               filter(days_apart <= 100,
                      days_apart >= 10))

( bac_fitted.emt <- emtrends(bac_t1d_lm, "T1D", var = "bray_curtis") )

pw_bac_t1d <- pairs(bac_fitted.emt)

pw_bac_t1d
```


```{r}
bac_days_p <- bac_next_bray_dt %>%
  filter(days_apart <= 100,
         days_apart >=10) %>%
  ggplot(aes(x= days_apart, y = bray_curtis, color = T1D)) +
  geom_point(alpha = 0.1) +
  geom_smooth(
    method = "lm"
    ) +
  geom_text(y = 0.9, x = 50,
            label = "p-value = 0.0133",
            color = "black") +
  scale_color_manual(values = c("#8CBEB1", "orangered")) +
  theme_bw() +
  labs(x = "Days to Next Sample",
       y = "Bray-Curtis Dissimilarity")

bac_days_p

ggsave(
  bac_days_p,
  file = sprintf(
    "%s/charts/bray_curtis_next_bac_days1.pdf",
    find_rstudio_root_file()
  ),
  height = 5, width = 3.5
)
```

average bray distance by subject/pair BACTERIOME
```{r}
bac_bray_sum_subj <- bac_next_bray_dt %>%
  group_by(mask_id1) %>%
  summarize(average_bray = mean(bray_curtis),
            T1D = T1D,
            pair_num = case_ind) %>%
  ungroup() %>%
  distinct() %>%
  filter(!pair_num == 114)
```

```{r}


bac_bray_subj_wide <- bac_bray_sum_subj %>%
  group_by(pair_num) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  pivot_wider(id_cols = pair_num, names_from = T1D, values_from = average_bray)
```

```{r}
bac_bray_subj_wide %>%
  ggpaired(cond1 = "Yes", cond2 = "No",
    fill = "condition", palette = c("#8CBEB1", "orangered"),
    line.color = "lightgrey", ggtheme = theme_bw(), line.size = 0.1) +
  stat_compare_means(paired = TRUE) 
  
```

```{r}
wilcox.test(bac_bray_subj_wide$Yes, bac_bray_subj_wide$No, paired = TRUE)

```

new plot average bray distance by T1D and country status VIROME
```{r}

bac_bray_sum_subj %>%
  group_by(pair_num) %>%
  filter(n() == 2) %>%
  ggplot(aes(T1D, average_bray)) +
  geom_boxplot(aes(fill = T1D), outlier.shape = NA, alpha = 0.6) +
  geom_point(color="grey20", alpha = 0.5, stroke = 0) +
  geom_line(aes(group = pair_num), color = "lightgrey", alpha = 0.5) +
  geom_text(label = "Wilcoxon,\np-value = 0.0009", x = "No", 
            y = 1, size = 8/.pt) +
  scale_fill_manual(values=c("#8CBEB1", "orangered")) +
  ylim(c(0.2, 1.05)) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_blank()) +
  labs(y = "Bray-Curtis Distance\nto Follow-up Sample",
       title="bacteria SGBs")

ggsave(file = sprintf("%s/charts/bacteriome_bray_curtis_T1D_NCC_paired.pdf",
                      find_rstudio_root_file()), 
                 width = 2.5, height = 3.5)  
```


DOL BACTERIOME
```{r}
## this functions rounds a number up to the nearest 100
round_any = function(x, accuracy, f=ceiling){f(x/ accuracy) * accuracy}

DOL_bac_pair_dt <- bac_next_bray_dt %>%
  mutate(rounded_DOL =  round_any(age_days, 100)) %>%
  filter(rounded_DOL <= 1400) %>%
  group_by(mask_id1, rounded_DOL) %>%
  summarize(average_bray = mean(bray_curtis),
            T1D = T1D,
            pair_num = case_ind) %>%
  ungroup() %>%
  distinct() %>%
  filter(!pair_num == 114)
```

```{r}
DOL_bac_pair_subj_wide <- DOL_bac_pair_dt %>%
  group_by(pair_num, rounded_DOL) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(pair_num, rounded_DOL), names_from = T1D, values_from = average_bray)
```

```{r}
DOL_bac_pair_subj_wide %>%
  ggpaired(cond1 = "Yes", cond2 = "No",
    fill = "condition", palette = c("#8CBEB1", "orangered"),
    line.color = "lightgrey", ggtheme = theme_bw(), line.size = 0.1,
    facet.by = "rounded_DOL") +
  stat_compare_means(paired = TRUE) 
```

```{r}
DOL_bac_pair_wilcox_dt <- DOL_bac_pair_subj_wide %>% 
  group_by(rounded_DOL) %>%
  do(w = wilcox.test(.$Yes, .$No, data=., paired=TRUE)) %>% 
       summarise(rounded_DOL, 
                 Wilcox = w$p.value,
                 aster = case_when(Wilcox < 0.0001 ~ "****",
                                   Wilcox < 0.001 ~ "***", 
                                   Wilcox < 0.01 ~ "**",
                                   Wilcox < 0.05 ~ "*",
                                   TRUE ~ "ns"))

DOL_bac_pair_wil_dt <- merge(DOL_bac_pair_dt, DOL_bac_pair_wilcox_dt, by = "rounded_DOL")
```

```{r}
DOL_bac_pair_wil_dt %>%
  group_by(rounded_DOL,pair_num) %>%
  filter(n() == 2) %>%
  ggplot(aes(T1D, average_bray)) +
  geom_boxplot(aes(fill = T1D), outlier.shape = NA, alpha = 0.6) +
  geom_point(color="grey20", alpha = 0.5, stroke = 0, size = 0.8) +
  geom_line(aes(group = pair_num), color = "lightgrey", alpha = 0.3) +
  geom_text(data = DOL_bac_pair_wilcox_dt, aes(label = aster), x = "No", y = 1.05) +
  scale_fill_manual(values=c("#8CBEB1", "orangered")) +
  ylim(c(NA, 1.1)) +
  facet_wrap(vars(rounded_DOL), nrow = 1) +
  theme_bw() +
  theme(legend.position = "bottom", 
        axis.text.x = element_blank()) +
  labs(x = "day of life (rounded)", 
       y = "Bray-Curtis Distance\nto Follow-up Sample",
       title="bacteria SGBs")

ggsave(file = sprintf("%s/charts/bacteriome_T1D_and_dayoflife_bray_curtis_NCC_paired.pdf",
                      find_rstudio_root_file()), 
                 width = 6, height = 3.5)  
```









