---
title: "plot abundance of important features from random forest COUNTRY"
output: html_notebook
---

load packages
```{r}
library(ggplot2)
library(data.table)
library(stringr)
library(dplyr)
library(tibble)
library(tidyr)
library(rprojroot)
library(cowplot)
library(latrend)
library(nationalparkcolors)

```

set paths and filenames
```{r}
### files
long_table=sprintf("%s/data/mp142_TGVG1.1_MPA4_combined_abundance_table_longform1.tsv", find_rstudio_root_file())
metadata_table=sprintf("%s/data/some_teddy_MP142_metadata2.all_samples1.delivery.csv", find_rstudio_root_file())

swe_important=sprintf("%s/intermediate_files/random_forest_outputs/rf_important_all_sp_Sweden_vs_not1_pat_groups.tsv", find_rstudio_root_file())
fin_important=sprintf("%s/intermediate_files/random_forest_outputs/rf_important_all_sp_Finland_vs_not1_pat_groups.tsv", find_rstudio_root_file())
ger_important=sprintf("%s/intermediate_files/random_forest_outputs/rf_important_all_sp_Germany_vs_not1_pat_groups.tsv", find_rstudio_root_file())
usa_important=sprintf("%s/intermediate_files/random_forest_outputs/rf_important_all_sp_USA_vs_not1_pat_groups.tsv", find_rstudio_root_file())

```

load long abundance table and metadata table, group abundance by SGBs
```{r}
long_dt <- fread(sprintf("%s", long_table), sep = "\t", header = T) %>%
  mutate(species = gsub(".*s__","s__", lineage ), 
         genus = gsub(".*g__","g__", lineage ),
         sampleID = as.character(sampleID)) %>%
  mutate(genus = gsub("\\|s__.*", "", genus),
         kingdom = case_when(grepl("k__Bac", lineage) ~ "Bacteria", 
                           grepl("k__Vir", lineage) ~ "Virus",
                           grepl("k__Ar", lineage) ~ "Archea",
                           grepl("k__Euk", lineage) ~ "Eukaryota",
                           TRUE ~ "other")) %>%
  subset(select = c("rel_abundance", "sampleID", "species", "genus", "kingdom"))


meta_dt <- fread(sprintf("%s", metadata_table), sep = ",", header = T) %>%
  select(-V1) %>%
  mutate(Country = case_when(
           country == 1 ~ "USA",
           country == 2 ~ "FIN",
           country == 3 ~ "GER",
           country == 4 ~ "SWE",
           TRUE ~ "other"
         ),
         sample = as.character(sample))

## this functions rounds a number up to the nearest 100
round_any = function(x, accuracy, f=ceiling){f(x/ accuracy) * accuracy}

```

get Sweden top 12 important features
```{r}
SWE_imp_dt <- fread(sprintf("%s", swe_important), sep = "\t", header = T) %>%
  head(n = 12) %>%
  select(-avg_importance)

SWE_ab_dt <- merge(long_dt, SWE_imp_dt, by = "species") 

SWE_ab_meta_dt <- merge(SWE_ab_dt, meta_dt, by.x = "sampleID", by.y = "sample") %>%
  mutate(rounded_DOL =  round_any(age_days, 100)) %>%
  filter(rounded_DOL <= 1400) %>%
  select(c(sampleID, species, rel_abundance, rounded_DOL, Country)) %>%
  group_by(species, rounded_DOL, Country) %>%
  ## relative abundance to itself over all samples from each time bracket
  summarize(avg_rel_abund = mean(rel_abundance)) %>%
  ungroup() %>%
  ## adding 0s for rounded_DOL where species was never detected
  complete(species, rounded_DOL, Country) %>%
  replace(is.na(.), 0)

```

plot Sweden top 12 important features longitudinally
```{r}
SWE_ab_meta_dt %>%
  mutate(species = gsub("s__", "", species)) %>%
  mutate(species = gsub("\\|t__", " ", species)) %>%
  mutate(species = gsub("_", " ", species)) %>%
  ggplot(aes(x = rounded_DOL, y = avg_rel_abund, color = Country)) +
  geom_line() +
  facet_wrap(vars(species), scales = "free_y", 
             labeller = label_wrap_gen()) +
  theme_cowplot() + 
  labs(y="average abundance", x="day of life", title="SWE important features") +
  scale_color_manual(values=c("#eddccb", "#8D7F99", "#163343", "#EC8FA3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        strip.text.x = element_text(size=6))

ggsave(file = sprintf("%s/charts/all_SGBs_SWE_rf_top12_features.pdf", find_rstudio_root_file()), 
                 width = 8, height = 4)
```

get Finland top 12 important features
```{r}
FIN_imp_dt <- fread(sprintf("%s", fin_important), sep = "\t", header = T) %>%
  head(n = 12) %>%
  select(-avg_importance)

FIN_ab_dt <- merge(long_dt, FIN_imp_dt, by = "species") 

FIN_ab_meta_dt <- merge(FIN_ab_dt, meta_dt, by.x = "sampleID", by.y = "sample") %>%
  mutate(rounded_DOL =  round_any(age_days, 100)) %>%
  filter(rounded_DOL <= 1400) %>%
  select(c(sampleID, species, rel_abundance, rounded_DOL, Country)) %>%
  group_by(species, rounded_DOL, Country) %>%
  ## relative abundance to itself over all samples from each time bracket
  summarize(avg_rel_abund = mean(rel_abundance)) %>%
  ungroup() %>%
  ## adding 0s for rounded_DOL where species was never detected
  complete(species, rounded_DOL, Country) %>%
  replace(is.na(.), 0)

```

plot Finland top 12 important features longitudinally
```{r}
FIN_ab_meta_dt %>%
  mutate(species = gsub("s__", "", species)) %>%
  mutate(species = gsub("\\|t__", " ", species)) %>%
  mutate(species = gsub("_", " ", species)) %>%
  ggplot(aes(x = rounded_DOL, y = avg_rel_abund, color = Country)) +
  geom_line() +
  facet_wrap(vars(species), scales = "free_y", 
             labeller = label_wrap_gen()) +
  theme_cowplot() + 
  labs(y="average abundance", x="day of life", title="FIN important features") +
  scale_color_manual(values=c("#eddccb", "#8D7F99", "#163343", "#EC8FA3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        strip.text.x = element_text(size=6))

ggsave(file = sprintf("%s/charts/all_SGBs_FIN_rf_top12_features.pdf", find_rstudio_root_file()), 
                 width = 8, height = 4)
```

get Germany top 12 important features
```{r}
GER_imp_dt <- fread(sprintf("%s", ger_important), sep = "\t", header = T) %>%
  head(n = 12) %>%
  select(-avg_importance)

GER_ab_dt <- merge(long_dt, GER_imp_dt, by = "species") 

GER_ab_meta_dt <- merge(GER_ab_dt, meta_dt, by.x = "sampleID", by.y = "sample") %>%
  mutate(rounded_DOL =  round_any(age_days, 100)) %>%
  filter(rounded_DOL <= 1400) %>%
  select(c(sampleID, species, rel_abundance, rounded_DOL, Country)) %>%
  group_by(species, rounded_DOL, Country) %>%
  ## relative abundance to itself over all samples from each time bracket
  summarize(avg_rel_abund = mean(rel_abundance)) %>%
  ungroup() %>%
  ## adding 0s for rounded_DOL where species was never detected
  complete(species, rounded_DOL, Country) %>%
  replace(is.na(.), 0)

```

plot Germany top 12 important features longitudinally
```{r}
GER_ab_meta_dt %>%
  mutate(species = gsub("s__", "", species)) %>%
  mutate(species = gsub("\\|t__", " ", species)) %>%
  mutate(species = gsub("_", " ", species)) %>%
  ggplot(aes(x = rounded_DOL, y = avg_rel_abund, color = Country)) +
  geom_line() +
  facet_wrap(vars(species), scales = "free_y", 
             labeller = label_wrap_gen()) +
  theme_cowplot() + 
  labs(y="average abundance", x="day of life", title="GER important features") +
  scale_color_manual(values=c("#eddccb", "#8D7F99", "#163343", "#EC8FA3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        strip.text.x = element_text(size=6))

ggsave(file = sprintf("%s/charts/all_SGBs_GER_rf_top12_features.pdf", find_rstudio_root_file()), 
                 width = 8, height = 4)
```

get USA top 12 important features
```{r}
USA_imp_dt <- fread(sprintf("%s", usa_important), sep = "\t", header = T) %>%
  head(n = 12) %>%
  select(-avg_importance)

USA_ab_dt <- merge(long_dt, USA_imp_dt, by = "species") 

USA_ab_meta_dt <- merge(USA_ab_dt, meta_dt, by.x = "sampleID", by.y = "sample") %>%
  mutate(rounded_DOL =  round_any(age_days, 100)) %>%
  filter(rounded_DOL <= 1400) %>%
  select(c(sampleID, species, rel_abundance, rounded_DOL, Country)) %>%
  group_by(species, rounded_DOL, Country) %>%
  ## relative abundance to itself over all samples from each time bracket
  summarize(avg_rel_abund = mean(rel_abundance)) %>%
  ungroup() %>%
  ## adding 0s for rounded_DOL where species was never detected
  complete(species, rounded_DOL, Country) %>%
  replace(is.na(.), 0)

```

plot USA top 12 important features longitudinally
```{r}
USA_ab_meta_dt %>%
  mutate(species = gsub("s__", "", species)) %>%
  mutate(species = gsub("\\|t__", " ", species)) %>%
  mutate(species = gsub("_", " ", species)) %>%
  ggplot(aes(x = rounded_DOL, y = avg_rel_abund, color = Country)) +
  geom_line() +
  facet_wrap(vars(species), scales = "free_y", 
             labeller = label_wrap_gen()) +
  theme_cowplot() + 
  labs(y="average abundance", x="day of life", title="USA important features") +
  scale_color_manual(values=c("#eddccb", "#8D7F99", "#163343", "#EC8FA3")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), 
        strip.text.x = element_text(size=6))

ggsave(file = sprintf("%s/charts/all_SGBs_USA_rf_top12_features.pdf", find_rstudio_root_file()), 
                 width = 8, height = 4)
```




